### 2012-03-14 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\begin{algorithm}
\caption{Function \texttt{getFunctionDeclarationNodes}: Retrieval of Function Declaration Nodes}
\label{ALG_retrieval_of_function_declaration_nodes}
\begin{algorithmic}[1]
\REQUIRE syntaxTree: A reference to the root node of a syntax tree.
\ENSURE A list of syntax tree nodes representing function declarations.

\STATE prologNode $=$ null
\FORALL{moduleChildNode $\in$ syntaxTree.getChildren()}
    \IF{is(moduleChildNode, PrologNode)}
        \STATE prologNode $=$ moduleChildNode
    \ENDIF
\ENDFOR

\STATE functionDeclarationNodes $=$ an empty list
\IF{prologNode $\neq$ null}
    \FORALL{prologChildNode $\in$ prologNode.getChildren()}
        \IF{is(prologChildNode, FunctionDeclNode)}
            \STATE add prologChildNode to functionDeclarationNodes
        \ENDIF
    \ENDFOR
\ENDIF

@@

\begin{algorithm}
\caption{Processing of Functions}
\label{ALG_processing_of_functions}
\begin{algorithmic}[1]
\REQUIRE syntaxTree: A reference to the root node of a syntax tree.
\ENSURE An associative array of function names with their types and references to the their declaration nodes.

\STATE functionArray $=$ an empty associative array
\FORALL{functionDeclarationNode $\in$ \\ getFunctionDeclarationNodes(syntaxTree)}
    \STATE functionName $=$ functionDeclarationNode.funcName
    \STATE typeNode $=$ functionDeclarationNode.getChild(TypeNode)
    \STATE functionArray[functionName] $=$ \\ (getTypeTN(typeNode), functionDeclarationNode)
\ENDFOR
\RETURN functionArray
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{getTypeTN}: Extraction of a Type from TypeNode}
\label{ALG_extraction_of_a_type_from_typenode}
\begin{algorithmic}[1]
\REQUIRE typeNode: Syntax tree node of TypeNode class.
\ENSURE Type extracted from typeNode.

\STATE type = null
\STATE cardinality = typeNode.cardinality
\STATE itemTypeNode = typeNode.getChild(ItemTypeNode)
\IF{is(itemTypeNode, AtomicTypeNode)}
    \STATE type = itemTypeNode.typeName
\ELSIF{is(itemTypeNode, KindTestNode)}
    \STATE type = itemTypeNode.nodeKind
\ENDIF

\RETURN type, cardinality
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{determineExpressionType}}
\label{ALG_func_determineExpressionType}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    exprNode: An expression node in the syntax tree. \\
    contextVarTypes: Types of local variables valid in the current subtree.
}
\ENSURE The expression type.

\STATE type $=$ UnknownType
\IF{is(exprNode, LiteralNode)}
     \STATE type $=$ XSD atomic type exprNode.type, exactly-once cardinality
\ELSIF{is(exprNode, FunctionCallNode)}
     \STATE type $=$ getFunctionReturnType(exprNode.fncName)
\ELSIF{is(exprNode, VarRefNode)}
     \STATE type = getVariableType(contextVarTypes, exprNode.varName)
     \IF{type is PathType}
         \STATE step $=$ a path step (new instance of StepExprNode) representing the variable reference
		 \STATE substeps = empty associative array         
         \STATE substeps[step] $=$ type
         \STATE pathExprNode $=$ new instance of PathExprNode containing one step step
         \STATE type $=$ PathType containing pathExprNode and substeps
     \ENDIF
\ELSIF{is(exprNode, PathExprNode)}
     \STATE type $=$ createPathType(exprNode)
\ELSIF{is(exprNode, OperatorNode)}
     \STATE type $=$ determineOperatorType(exprNode, contextVarTypes)
\ELSIF{is(exprNode, FLWORExprNode)}
     \STATE returnClauseNode $=$ exprNode.getChild(ReturnClauseNode)
     \STATE type $=$ createForUnboundType(getType(returnClauseNode))
\ENDIF

@@

\STATE $type =$ UnknownType
\STATE $operator =$ $operatorNode.operator$
\IF{isOperatorClassComparison($operator$)}
    \STATE $type =$ boolean, exactly-one cardinality
\ELSIF{isOperatorClassAddition($operator$)}
    \STATE $leftOperandType =$ getType($exprNode.leftSide$)
    \STATE $rightOperandType =$ getType($exprNode.rightSide$)
    \IF{isNumericType($leftOperandType$) $\wedge$ isNumericType($rightOperandType$)}
        \STATE $type =$ selectMoreGeneralNumericType($leftOperandType$, $rightOperandType$)
    \ELSIF{isNumericType($leftOperandType$)}
        \STATE $type = leftOperandType$
    \ELSIF{isNumericType($rightOperandType$)}
        \STATE $type = rightOperandType$
    \ENDIF
\ENDIF

@@

\begin{algorithm}
\caption{Function \texttt{isOperatorClassComparison}}
\label{ALG_func_isOperatorClassComparison}
\begin{algorithmic}[1]
\REQUIRE $operator$: Representation of a XQuery operator.
\ENSURE true if operator is a comparison operator, false otherwise.

@@

\begin{algorithm}
\caption{Function isOperatorClassAddition()}
\label{ALG_func_isOperatorClassAddition}
\begin{algorithmic}[1]
\REQUIRE $operator$: Representation of a XQuery operator.
\ENSURE $true$ if $operator$ is an addition operator, $false$ otherwise.

\IF{$operator$ equals one of $PLUS$, $MINUS$, $UNARY\_PLUS$, $UNARY\_MINUS$}
    \RETURN $true$
\ELSE
    \RETURN $false$
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function isNumericType()}
\label{ALG_func_isNumericType}
\begin{algorithmic}[1]
\REQUIRE $type$: Representation of a type.
\ENSURE $true$ if $type$ represents one of the XSD atomic built-in numeric types, $false$ otherwise.

\IF{$type$ represents one of $float$, $double$, $decimal$, $integer$, $long$, $int$, $short$, $byte$, $nonPositiveInteger$, $negativeInteger$, $nonNegativeInteger$, $positiveInteger$, $unsignedLong$, $unsignedInt$, $unsignedShort$, $unsignedByte$}
    \RETURN $true$
\ELSE
    \RETURN $false$
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function analysisOfExpressionTypes}
\label{ALG_func_analysisOfExpressionTypes}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $startingNode$: A node determining a subtree to perform the analysis on. \\
    $contextVarTypes$: Types of local variables valid in the current subtree.
}

\FORALL{$i \in \{1, \dots, |startingNode.getChildren()|\}$ ordered from the lowest to the highest}
    \STATE $subnode = startingNode.getChild(i)$
    \STATE $analysisOfExpressionTypes(subnode, contextVarTypes)$ \COMMENT{recursion}
    \STATE $newVars = get(subnode, newVars)$    
    \STATE $contextVarTypes = mergeContextVarTypes(contextVarTypes, newVars)$
\ENDFOR
\IF{$is(startingNode, ExprNode)$}
    \STATE $memorizeType(startingNode, determineExpressionType(startingNode))$
\ENDIF
\IF{$is(startingNode, VariableBindingNode)$}
    \STATE $type = null$
    \STATE $typeNode = startingNode.getChild(TypeNode)$
    \IF{$typeNode \neq null$}
        \STATE $type = getTypeTN(typeNode)$
    \ELSE
        \STATE $type = determineExpressionType(startingNode.$ $getChild(BindingSequenceNode).getChild(ExprNode),$ $contextVarTypes)$
        \IF{$is(startingNode, ForClauseNode)$}
            \STATE $type = createForBoundType(type)$
        \ENDIF
    \ENDIF
    \STATE $set(startingNode, newVars, (startingNode.varName, type))$
\ELSIF{$is(startingNode, TupleStreamNode)$}
    \FORALL{$varBindingNode \in startingNode.getChildren()$}
        \STATE $add(startingNode, newVars, get(varBindingNode, newVars))$
    \ENDFOR
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function mergeContextVarTypes}
\label{ALG_function_mergeContextVarTypes}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $keys()$: Function retrieving index values of an indexed structure. \\
    $contextVarTypes$: Types of local variables valid in current context. \\    
    $extendingVarTypes$: New variables in the same structure as the first argument.
}
\ENSURE The context variable types extended with the variable types from the second argument.

\FORALL{$varName \in keys(extendingVarTypes)$}
    \STATE $contextVarTypes[varName] = extendingVarTypes[varName]$
\ENDFOR
\RETURN $contextVarTypes$
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function getVariableType}
\label{ALG_function_get_variable_type}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $contextVarTypes$: Types of local variables valid in current context. \\
    $var$: A variable to determine the type of.
}
\ENSURE The type of variable $var$.

@@

\begin{algorithm}
\caption{Function createPathType}
\label{ALG_function_create_path_type}
\begin{algorithmic}[1]
\REQUIRE $pathExprNode$: A reference to a PathExprNode to create the PathType from.
\ENSURE The PathType of $pathExprNode$ expression.

\STATE $stepNodes =$ an empty array
\STATE $forBoundSubsteps =$ an empty associative array
\FORALL{$stepNode \in pathExprNode.getChildren()$}
	\STATE $detailNode = stepNode.getChild(ExprNode)$
	\IF{$detailNode != null$}
		\IF{$is(detailNode, VarRefNode)$}
			\STATE $type = getType(detailNode)$ // $PathType$
			\IF{$type.isForBound$}
				\STATE $forBoundSubsteps[stepNode] = type$
				\STATE add $stepNode$ to $stepNodes$
			\ELSE
				\STATE add all $type.stepNodes$ to $stepNodes$
			\ENDIF
		\ELSE
			\STATE add $stepNode$ to $stepNodes$
		\ENDIF
	\ELSE
		\STATE add $stepNode$ to $stepNodes$
	\ENDIF
\ENDFOR

\RETURN $PathType$ with $stepNodes$, $forBoundSubsteps$, and $isForBound$ set to $false$
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function createForBoundType}
\label{ALG_function_create_for_bound_type}
\begin{algorithmic}[1]
\REQUIRE $type$: A type to create the for bound type from.
\ENSURE The for bound type from $type$.

\IF{$type$ is $UnknownType$}
	\RETURN $UnknownType$
\ELSIF{$type$ is $NodeType$ or XSD atomic type}
	\RETURN $type$ with cardinality set to exactly-one
\ELSE
	\STATE // $PathType$
	\RETURN $type$ with $isForBound$ flag set to $true$
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function createForUnboundType}
\label{ALG_function_create_for_unbound_type}
\begin{algorithmic}[1]
\REQUIRE $type$: A type to create the for unbound type from.
\ENSURE The for unbound type from $type$.

\IF{$type$ is $UnknownType$}
	\RETURN $UnknownType$
\ELSIF{$type$ is $NodeType$ or XSD atomic type}
	\RETURN $type$ with cardinality set to zero-or-more
\ELSE
	\STATE // $PathType$
	\RETURN $type$ with $isForBound$ flag set to $false$
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function usesOnlyChildAndDescendantAxes}
\label{ALG_function_uses_only_child_and_descendant_axes}
\begin{algorithmic}[1]
\REQUIRE $pathType$: A $PathType$ instance.
\ENSURE A boolean result.

\FORALL{$stepNode \in pathType.stepNodes$}
	\IF{$stepNode.isAxisStep$}
		\STATE $axisKind$ = $stepNode.axisNode.getAxisKind()$
		\IF{$axisKind$ $\neq$ $CHILD$ AND $axisKind$ $\neq$ $DESCENDANT$}
			\RETURN false
		\ENDIF
	\ENDIF
\ENDFOR

\STATE $detailNode$ = $stepNode.getChild(ExprNode)$
\IF{$detailNode \neq$ null}
	\IF{$is(detailNode, VarRefNode)$}
		\IF{$usesOnlyChildAndDescendantAxes($ $pathType.forBoundSubsteps[stepNode])$ == false}
			\RETURN false
		\ENDIF
	\ENDIF
\ENDIF

@@

\begin{algorithm}
\caption{Function isWithoutPredicates}
\label{ALG_function_is_without_predicates}
\begin{algorithmic}[1]
\REQUIRE $pathType$: A $PathType$ instance.
\ENSURE A boolean result.getTargetReturnPathTypesRecursive

\FORALL{$stepNode \in pathType.stepNodes$}
	\IF{$stepNode.hasPredicates()$}
		\RETURN false
	\ENDIF
	\IF{$pathType.forBoundSubsteps[stepNode]$ $\neq$ null}
		\IF{$isWithoutPredicates(pathType.forBoundSubsteps[stepNode]$ == false}
			\RETURN false
		\ENDIF
	\ENDIF
\ENDFOR

@@

\begin{algorithm}
\caption{Function endsWithExactlyOnePredicate}
\label{ALG_function_ends_with_exactly_one_predicate}
\begin{algorithmic}[1]
\REQUIRE $pathType$: A $PathType$ instance.
\ENSURE If $pathType$ ends with exactly one predicate, return value is the predicate, otherwise return value is false.

\STATE $lastStepNode$ = last item from $pathType.stepNodes$
\IF{$lastStepNode.hasPredicates()$ $==$ false}
	\RETURN false
\ENDIF
\IF{number of item in list $lastStepNode.getChild(PredicateListNode)$ $.getChildren()$ is higher than 1}
	\RETURN false
\ENDIF

@@

\begin{algorithm}
\caption{Function isTargetPath}
\label{ALG_function_is_target_path}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $pathType$: A $PathType$ instance. \\
  $varName$: A variable name.
}
\ENSURE Returns true, if the first step of path represented by the $PathType$ instance is a reference to the specified variable.

\todo[inline]{Upravit styl vsetkych algoritmov (null, ==, !=, true/false, ...) tak ako je tu.}

\STATE $firstStep$ = first item from  $pathType.stepNodes$
\STATE $detailNode$ = $firstStep.detailNode$
\IF{$detailNode$ == null}
	\RETURN false
\ENDIF
\IF{$is(detailNode, VarRefNode)$ == false}
	\RETURN false
\ENDIF
\IF{$detailNode.varName$ == $varName$}
	\RETURN true
\ELSE
	\RETURN false
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function getTargetReturnPathTypes}
\label{ALG_function_get_target_return_path_types}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $flworNode$: A $FLWORExprNode$ instance to search for target return paths. \\
  $varName$: A variable name.
}
\ENSURE A list of target return paths for the specified variable, in the given FLWOR, represented by $PathType$ instances.

\STATE $pathTypes$ = an empty array
\STATE $getTargetReturnPathTypesRecursive(flworNode.$ $getChild(ReturnClauseNode), varName, pathTypes)$
\RETURN $pathTypes$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function getTargetReturnPathTypesRecursive}
\label{ALG_function_get_target_return_path_types_recursive}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $node$: A node to search for target return paths. \\
  $varName$: A variable name. \\
  $pathTypes$ An output list to add found target return path types to.
}

\IF{$is(node, ExprNode)$}
	\STATE $type$ = $getType(node)$
	\IF{$type$ is $PathType$ AND $isTargetPath(type, varName)$}
		\STATE add $type$ to $pathTypes$
	\ENDIF
\ENDIF
\FORALL{$child \in node.getChildren()$}
	\STATE $getTargetReturnPathTypesRecursive(child, varName, pathTypes)$
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function getTargetReturnPaths}
\label{ALG_function_get_target_return_paths}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $flworNode$: A $FLWORExprNode$ instance to search for target return paths. \\
  $varName$: A variable name.
}
\ENSURE A list of target return paths for the specified variable, in the given FLWOR, represented by $PartExprNode$ instances.

\STATE $paths$ = an empty array
\STATE $getTargetReturnPathsRecursive(flworNode.$ $getChild(ReturnClauseNode), varName, paths)$
\RETURN $paths$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function getTargetReturnPathsRecursive}
\label{ALG_function_get_target_return_paths_recursive}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $node$: A node to search for target return paths. \\
  $varName$: A variable name. \\
  $paths$ An output list to add found target return path types to.
}

>>>>

\begin{algorithm}
\caption{Function \texttt{getFunctionDeclarationNodes}: Retrieval of Function Declaration Nodes}
\label{ALG_retrieval_of_function_declaration_nodes}
\begin{algorithmic}[1]
\REQUIRE $syntaxTree$: A reference to the root node of a syntax tree.
\ENSURE A list of syntax tree nodes representing function declarations.

\STATE $prologNode :=$ null
\FORALL{$moduleChildNode \in syntaxTree.getChildren()$}
    \IF{$is(moduleChildNode,$ PrologNode$)$}
        \STATE $prologNode := moduleChildNode$
    \ENDIF
\ENDFOR

\STATE $functionDeclarationNodes :=$ an empty list
\IF{$prologNode \neq$ null}
    \FORALL{$prologChildNode \in prologNode.getChildren()$}
        \IF{$is(prologChildNode,$ FunctionDeclNode$)$}
            \STATE add $prologChildNode$ to $functionDeclarationNodes$
        \ENDIF
    \ENDFOR
\ENDIF

@@

\begin{algorithm}
\caption{Processing of Functions}
\label{ALG_processing_of_functions}
\begin{algorithmic}[1]
\REQUIRE $syntaxTree$: A reference to the root node of a syntax tree.
\ENSURE An associative array of function names with their types and references to the their declaration nodes.

\STATE $functionArray :=$ an empty associative array
\FORALL{$functionDeclarationNode \in$ \\ $getFunctionDeclarationNodes(syntaxTree)$}
    \STATE $functionName := functionDeclarationNode.funcName$
    \STATE $typeNode := functionDeclarationNode.getChild(TypeNode)$
    \STATE $functionArray[functionName] :=$ \\ $(getTypeTN(typeNode), functionDeclarationNode)$
\ENDFOR
\RETURN $functionArray$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{getTypeTN}: Extraction of a Type from TypeNode}
\label{ALG_extraction_of_a_type_from_typenode}
\begin{algorithmic}[1]
\REQUIRE $typeNode$: Syntax tree node of TypeNode class.
\ENSURE Type extracted from $typeNode$.

\STATE $type :=$ null
\STATE $cardinality := typeNode.cardinality$
\STATE $itemTypeNode := typeNode.getChild(ItemTypeNode)$
\IF{$is(itemTypeNode,$ AtomicTypeNode$)$}
    \STATE $type := itemTypeNode.typeName$
\ELSIF{$is(itemTypeNode,$ KindTestNode$)$}
    \STATE $type := itemTypeNode.nodeKind$
\ENDIF

\RETURN $type, cardinality$
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{determineExpressionType}}
\label{ALG_func_determineExpressionType}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $exprNode$: An expression node in the syntax tree. \\
    $contextVarTypes$: Types of local variables valid in the current subtree.
}
\ENSURE The expression type.

\STATE $type :=$ UnknownType
\IF{$is(exprNode,$ LiteralNode$)$}
     \STATE $type :=$ XSD atomic type $exprNode.type$, exactly-once cardinality
\ELSIF{$is(exprNode,$ FunctionCallNode$)$}
     \STATE $type := getFunctionReturnType(exprNode.fncName)$
\ELSIF{$is(exprNode,$ VarRefNode$)$}
     \STATE $type := getVariableType(contextVarTypes, exprNode.varName)$
     \IF{$type$ is PathType}
         \STATE $step :=$ a path step (new instance of StepExprNode) representing the variable reference
		 \STATE $substeps :=$ an empty associative array         
         \STATE $substeps[step] := type$
         \STATE $pathExprNode :=$ new instance of PathExprNode containing one step step
         \STATE $type :=$ PathType containing $pathExprNode$ and $substeps$
     \ENDIF
\ELSIF{$is(exprNode,$ PathExprNode$)$}
     \STATE $type := createPathType(exprNode)$
\ELSIF{$is(exprNode,$ OperatorNode$)$}
     \STATE $type := determineOperatorType(exprNode, contextVarTypes)$
\ELSIF{$is(exprNode,$ FLWORExprNode$)$}
     \STATE $returnClauseNode := exprNode.getChild(ReturnClauseNode)$
     \STATE $type := createForUnboundType(getType(returnClauseNode))$
\ENDIF

@@

\STATE $type :=$ UnknownType
\STATE $operator := operatorNode.operator$
\IF{$isOperatorClassComparison(operator)$}
    \STATE $type :=$ boolean, exactly-one cardinality
\ELSIF{$isOperatorClassAddition(operator)$}
    \STATE $leftOperandType := getType(exprNode.leftSide)$
    \STATE $rightOperandType := getType(exprNode.rightSide)$
    \IF{$isNumericType(leftOperandType)$ \\ $\wedge$ $isNumericType(rightOperandType)$}
        \STATE $type := selectMoreGeneralNumericType(leftOperandType,$ $rightOperandType)$
    \ELSIF{$isNumericType(leftOperandType)$}
        \STATE $type := leftOperandType$
    \ELSIF{$isNumericType(rightOperandType)$}
        \STATE $type := rightOperandType$
    \ENDIF
\ENDIF

@@

\begin{algorithm}
\caption{Function \texttt{isOperatorClassComparison}}
\label{ALG_func_isOperatorClassComparison}
\begin{algorithmic}[1]
\REQUIRE $operator$: A representation of a XQuery operator.
\ENSURE true if $operator$ is a comparison operator, false otherwise.

@@

\begin{algorithm}
\caption{Function \texttt{isOperatorClassAddition}}
\label{ALG_func_isOperatorClassAddition}
\begin{algorithmic}[1]
\REQUIRE $operator$: A representation of a XQuery operator.
\ENSURE true if $operator$ is an addition operator, false otherwise.

\IF{$operator$ equals one of PLUS, MINUS, UNARY\_PLUS, UNARY\_MINUS}
    \RETURN true
\ELSE
    \RETURN false
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{isNumericType}}
\label{ALG_func_isNumericType}
\begin{algorithmic}[1]
\REQUIRE $type$: A representation of a type.
\ENSURE true if $type$ represents one of the XSD atomic built-in numeric types, false otherwise.

\IF{$type$ represents one of float, double, decimal, integer, long, int, short, byte, nonPositiveInteger, negativeInteger, nonNegativeInteger, positiveInteger, unsignedLong, unsignedInt, unsignedShort, unsignedByte}
    \RETURN true
\ELSE
    \RETURN false
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{analysisOfExpressionTypes}}
\label{ALG_func_analysisOfExpressionTypes}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $startingNode$: A node determining a subtree to perform the analysis on. \\
    $contextVarTypes$: Types of local variables valid in the current subtree.
}

\FORALL{$i \in \{1, \dots, |startingNode.getChildren()|\}$ ordered from the lowest to the highest}
    \STATE $subnode := startingNode.getChild(i)$
    \STATE $analysisOfExpressionTypes(subnode, contextVarTypes)$ \COMMENT{recursion}
    \STATE $newVars := get(subnode, newVars)$    
    \STATE $contextVarTypes := mergeContextVarTypes(contextVarTypes, newVars)$
\ENDFOR
\IF{$is(startingNode,$ ExprNode$)$}
    \STATE $memorizeType(startingNode, determineExpressionType(startingNode))$
\ENDIF
\IF{$is(startingNode,$ VariableBindingNode$)$}
    \STATE $type :=$ null
    \STATE $typeNode := startingNode.getChild(TypeNode)$
    \IF{$typeNode \neq null$}
        \STATE $type := getTypeTN(typeNode)$
    \ELSE
        \STATE $type := determineExpressionType(startingNode.$ $getChild(BindingSequenceNode).getChild(ExprNode),$ $contextVarTypes)$
        \IF{$is(startingNode,$ ForClauseNode$)$}
            \STATE $type := createForBoundType(type)$
        \ENDIF
    \ENDIF
    \STATE $set(startingNode, newVars, (startingNode.varName, type))$
\ELSIF{$is(startingNode,$ TupleStreamNode$)$}
    \FORALL{$varBindingNode \in startingNode.getChildren()$}
        \STATE $add(startingNode, newVars, get(varBindingNode, newVars))$
    \ENDFOR
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{mergeContextVarTypes}}
\label{ALG_function_mergeContextVarTypes}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $contextVarTypes$: Types of local variables valid in current context. \\    
    $extendingVarTypes$: New variables in the same structure as the first argument.
}
\ENSURE The context variable types extended with the variable types from the second argument.

\FORALL{$varName \in keys(extendingVarTypes)$}
    \STATE $contextVarTypes[varName] := extendingVarTypes[varName]$
\ENDFOR
\RETURN $contextVarTypes$
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{getVariableType}}
\label{ALG_function_get_variable_type}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $contextVarTypes$: Types of local variables valid in current context. \\
    $var$: A variable to determine the type of.
}
\ENSURE The type of variable $var$.

@@

\begin{algorithm}
\caption{Function \texttt{createPathType}}
\label{ALG_function_create_path_type}
\begin{algorithmic}[1]
\REQUIRE $pathExprNode$: A reference to a PathExprNode to create the PathType from.
\ENSURE The PathType of $pathExprNode$ expression.

\STATE $stepNodes :=$ an empty array
\STATE $forBoundSubsteps :=$ an empty associative array
\FORALL{$stepNode \in pathExprNode.getChildren()$}
	\STATE $detailNode := stepNode.getChild(ExprNode)$
	\IF{$detailNode \neq$ null}
		\IF{$is(detailNode,$ VarRefNode$)$}
			\STATE $type := getType(detailNode)$ \COMMENT{PathType}
			\IF{$type.isForBound$}
				\STATE $forBoundSubsteps[stepNode] := type$
				\STATE add $stepNode$ to $stepNodes$
			\ELSE
				\STATE add all $type.stepNodes$ to $stepNodes$
			\ENDIF
		\ELSE
			\STATE add $stepNode$ to $stepNodes$
		\ENDIF
	\ELSE
		\STATE add $stepNode$ to $stepNodes$
	\ENDIF
\ENDFOR

\RETURN $PathType$ with $stepNodes$, $forBoundSubsteps$, and $isForBound$ set to false
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{createForBoundType}}
\label{ALG_function_create_for_bound_type}
\begin{algorithmic}[1]
\REQUIRE $type$: A type to create the for bound type from.
\ENSURE The for bound type from $type$.

\IF{$type$ is UnknownType}
	\RETURN UnknownType
\ELSIF{$type$ is NodeType or XSD atomic type}
	\RETURN $type$ with cardinality set to exactly-one
\ELSE
	\STATE \COMMENT {It is PathType}
	\RETURN $type$ with $isForBound$ flag set to true
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{createForUnboundType}}
\label{ALG_function_create_for_unbound_type}
\begin{algorithmic}[1]
\REQUIRE $type$: A type to create the for unbound type from.
\ENSURE The for unbound type from $type$.

\IF{$type$ is UnknownType}
	\RETURN UnknownType
\ELSIF{$type$ is NodeType or XSD atomic type}
	\RETURN $type$ with cardinality set to zero-or-more
\ELSE
	\STATE \COMMENT {It is PathType}
	\RETURN $type$ with $isForBound$ flag set to false
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{usesOnlyChildAndDescendantAxes}}
\label{ALG_function_uses_only_child_and_descendant_axes}
\begin{algorithmic}[1]
\REQUIRE $pathType$: A PathType instance.
\ENSURE A boolean result.

\FORALL{$stepNode \in pathType.stepNodes$}
	\IF{$stepNode.isAxisStep$}
		\STATE $axisKind := stepNode.axisNode.getAxisKind()$
		\IF{$axisKind \neq$ CHILD $\wedge$ $axisKind \neq$ DESCENDANT}
			\RETURN false
		\ENDIF
	\ENDIF
\ENDFOR

\STATE $detailNode := stepNode.getChild(ExprNode)$
\IF{$detailNode \neq$ null}
	\IF{$is(detailNode,$ VarRefNode$)$}
		\IF{$usesOnlyChildAndDescendantAxes($ $pathType.forBoundSubsteps[stepNode]) =$ false}
			\RETURN false
		\ENDIF
	\ENDIF
\ENDIF

@@

\begin{algorithm}
\caption{Function \texttt{isWithoutPredicates}}
\label{ALG_function_is_without_predicates}
\begin{algorithmic}[1]
\REQUIRE $pathType$: A PathType instance.
\ENSURE A boolean result.

\FORALL{$stepNode \in pathType.stepNodes$}
	\IF{$stepNode.hasPredicates$}
		\RETURN false
	\ENDIF
	\IF{$pathType.forBoundSubsteps[stepNode] \neq$ null}
		\IF{$isWithoutPredicates(pathType.forBoundSubsteps[stepNode] =$ false}
			\RETURN false
		\ENDIF
	\ENDIF
\ENDFOR

@@

\begin{algorithm}
\caption{Function \texttt{endsWithExactlyOnePredicate}}
\label{ALG_function_ends_with_exactly_one_predicate}
\begin{algorithmic}[1]
\REQUIRE $pathType$: A PathType instance.
\ENSURE If $pathType$ ends with exactly one predicate, return value is the predicate, otherwise return value is false.

\STATE $lastStepNode :=$ the last item from $pathType.stepNodes$
\IF{$lastStepNode.hasPredicates() =$ false}
	\RETURN false
\ENDIF
\IF{number of items in list $lastStepNode.getChild(PredicateListNode)$ $.getChildren()$ is higher than 1}
	\RETURN false
\ENDIF

@@

\begin{algorithm}
\caption{Function \texttt{isTargetPath}}
\label{ALG_function_is_target_path}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $pathType$: A PathType instance. \\
  $varName$: A variable name.
}
\ENSURE Returns true, if the first step of path represented by the PathType instance is a reference to the specified variable.

\STATE $firstStep :=$ first item from  $pathType.stepNodes$
\STATE $detailNode := firstStep.detailNode$
\IF{$detailNode =$ null}
	\RETURN false
\ENDIF
\IF{$is(detailNode, VarRefNode) =$ false}
	\RETURN false
\ENDIF
\IF{$detailNode.varName = varName$}
	\RETURN true
\ELSE
	\RETURN false
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{getTargetReturnPathTypes}}
\label{ALG_function_get_target_return_path_types}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $flworNode$: A FLWORExprNode instance to search for target return paths. \\
  $varName$: A variable name.
}
\ENSURE A list of target return paths for the specified variable, in the given FLWOR, represented by PathType instances.

\STATE $pathTypes :=$ an empty array
\STATE $getTargetReturnPathTypesRecursive(flworNode.$ $getChild(ReturnClauseNode), varName, pathTypes)$
\RETURN $pathTypes$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{getTargetReturnPathTypesRecursive}}
\label{ALG_function_get_target_return_path_types_recursive}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $node$: A node to search for target return paths. \\
  $varName$: A variable name. \\
  $pathTypes$ An output list to add found target return path types to.
}

\IF{$is(node,$ ExprNode$)$}
	\STATE $type := getType(node)$
	\IF{$type$ is $PathType$ $\wedge$ $isTargetPath(type, varName)$}
		\STATE add $type$ to $pathTypes$
	\ENDIF
\ENDIF
\FORALL{$child \in node.getChildren()$}
	\STATE $getTargetReturnPathTypesRecursive(child, varName, pathTypes)$
\ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function \texttt{getTargetReturnPaths}}
\label{ALG_function_get_target_return_paths}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $flworNode$: A FLWORExprNode instance to search for target return paths. \\
  $varName$: A variable name.
}
\ENSURE A list of target return paths for the specified variable, in the given FLWOR, represented by PathExprNode instances.

\STATE $paths :=$ an empty array
\STATE $getTargetReturnPathsRecursive(flworNode.$ $getChild(ReturnClauseNode), varName, paths)$
\RETURN $paths$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{getTargetReturnPathsRecursive}}
\label{ALG_function_get_target_return_paths_recursive}
\begin{algorithmic}[1]
\REQUIRE{\ \\
  $node$: A node to search for target return paths. \\
  $varName$: A variable name. \\
  $paths$ An output list to add found target return path types to.
}

\IF{$is(node,$ PathExprNode$)$}
	\STATE $type := getType(node)$
	\IF{$type$ is PathType $\wedge$ $isTargetPath(type, varName)$}
		\STATE add $type$ to $paths$
	\ENDIF
\ELSIF{$is(node,$ VarRefNode$)$}
	\IF{$node.varName = varName$}
		\STATE $type := getType(node)$
		\IF {$type$ is PathType $\wedge$ $isTargetPath(type, varName)$}
			\STATE add $type$ to $paths$
		\ENDIF
	\ENDIF
\ENDIF
\FORALL{$child \in node.getChildren()$}
	\STATE $getTargetReturnPathsRecursive(child, varName, paths)$
\ENDFOR
\end{algorithmic}
\end{algorithm}
